---
title: "NYPD Shooting Incident Data Report"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document: default
  html_document: default
theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

This assingment leverages data from NYPD Shooting Incident Data (Historic) (https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD) to explore NYPD Shooting Incident Data. This is a breakdown of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year. This data is manually extracted every quarter and reviewed by the Office of Management Analysis and Planning before being posted on the NYPD website. Each record represents a shooting incident in NYC and includes information about the event, the location and time of occurrence. In addition, information related to suspect and victim demographics is also included. We'll clean and analyze the data, using visualizations to uncover trends and insights.  This assingment involves importing the data, tidying it for analysis, and ensuring reproducibility through clear steps.  I will also consider potential biases within the data and draw conclusions based on our findings.


## Step 1: Import Library

```{r library, message=FALSE, warning=FALSE}
# Install required packages if not already installed
# install.packages("tidyverse")
# install.packages("lubridate")

library(tidyverse)
library(lubridate)
library(ggplot2)
```

## Step 2: Load Data

# Load the dataset from the provided URL

```{r load}
df_historic = read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
```

*Data Summary*

```{r}
# Show summary of dataframe
summary(df_historic)
```
```{r}
# Show  columns
colnames(df_historic)
```
## Step 3: Tidy and Transform Data

*Identify Relevant Columns*

Let's identify the columns relevant to analysis. I will focus on:

*OCCUR_DATE: Date of the incident
*OCCUR_TIME: Time of the incident
*BORO: Borough where the incident happened
*STATISTICAL_MURDER_FLAG: Indicates if the incident was a murder
*PERP_AGE_GROUP: Age group of the perpetrator
*PERP_SEX: Sex of the perpetrator
*PERP_RACE: Race of the perpetrator
*VIC_AGE_GROUP: Age group of the victim
*VIC_SEX: Sex of the victim
*VIC_RACE: Race of the victim

*Data Preprocessing*

```{r}
df_clean <- df_historic %>%
  select(c("OCCUR_DATE","OCCUR_TIME","BORO","PRECINCT",
           "STATISTICAL_MURDER_FLAG","VIC_AGE_GROUP","VIC_SEX","VIC_RACE")) %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE),
         OCCUR_TIME = hms(OCCUR_TIME),
         STATISTICAL_MURDER_FLAG = as.logical(STATISTICAL_MURDER_FLAG),
         Shootings = 1,
         Year = year(OCCUR_DATE))
colnames(df_clean)
print(df_clean)
```
 
*Address Outliers (Optional)**

Depending on the context, you might want to explore and handle outliers in specific variables.


**Explanation:**

*mdy(OCCUR_DATE): Converts the OCCUR_DATE (likely stored as a character string) into a date object using mdy(). This function assumes the format is month/day/year.
*wday(OCCUR_DAY, label = TRUE): Extracts the day of the week (e.g., "Monday", "Tuesday") from the date object OCCUR_DAY and stores it in a new variable. The label = TRUE argument ensures the output includes the day name.
*hour(hms(as.character(OCCUR_TIME))): Converts the OCCUR_TIME (likely stored as a character string) into a time object using hms(). Then, hour() extracts the hour (0-23) from the time object.
*as.factor(): Converts the following variables into factor data types:
*PERP_AGE_GROUP: Categorical data representing perpetrator age groups.
*PERP_SEX: Categorical data representing perpetrator sex.
*PERP_RACE: Categorical data representing perpetrator race.
*VIC_AGE_GROUP: Categorical data representing victim age groups.
*VIC_SEX: Categorical data representing victim sex.
*VIC_RACE: Categorical data representing victim race.
*BORO: Categorical data representing the borough where the incident occurred.

**Benefits of Converting Data Types:**

*Analysis:* Converting data types makes it easier to perform certain statistical analyses (e.g., calculating frequencies for categorical variables).
*Visualization:* Factor variables are often used to create informative visualizations like bar charts and boxplots.
Data Cleaning (Optional): Converting data types can sometimes help identify inconsistencies or errors in the data.

## Step 3: Add Visualizations and Analysis

*Research Question: Shooting areas and day analysis*

Question 1:  In which borough do most shooting incidents occur, and how do murder rates vary by day across boroughs?

Use: We'll use the dplyr package for data manipulation and ggplot2 for creating visualizations.

Summary: We'll analyze the number of incidents and calculate murder rates for each borough.

Visual:

Number of Incidents by Borough: Bar chart showing the distribution of incidents across boroughs.
Murder Rates by Borough: Bar chart or table showing the proportion of incidents classified as murders for each borough.

Conclusion: The borough with the most incidents might not necessarily have the highest murder rate.  Analyzing murder rates alongside raw incident counts provides a clearer picture of areas with potentially higher violent crime.

```{r}
# Incidents by Borough
ggplot(df_clean, aes(x = BORO)) +
  geom_bar(stat = "count") +
  labs(title = "Shooting Incidents by Borough",
       x = "Borough",
       y = "Count") +
  theme_minimal()
```


```{r}
df_clean %>%
  ggplot(aes(x = BORO, fill = BORO)) +
  geom_bar() +
  labs(title = "NYPD Shootings Incidents by Borough",
       subtitle = "(2006 - 2021)",
       x = "NYC Boroughs",
       y = "Total Number of Shootings")
```
*NYPD Shooting Incidents by Year:*

Type of Chart: Line chart with points
Analysis: This chart provides a different perspective on the yearly shooting incidents. It uses a line to connect the total number of shootings (Shootings) for each year (as a factor) and adds red points to highlight the data points. This visualization can help us identify trends and/or patterns in shooting incidents over time.

```{r}
df_year <- df_clean %>%
  group_by(Year) %>%
  summarize(Shootings = sum(Shootings))

df_year %>%
  ggplot(aes(x = as.factor(Year), y = Shootings)) +
  geom_line() +
  geom_point(color = "red") +
  scale_x_discrete(labels = as.character(2006:2021)) +
  labs(
    title = "NYPD Shooting Incidents by Year",
    x = "Years (2006 - 2021)",
    y = "Total Number of Shootings"
  )
```

*NYPD Shootings by Borough by Year:*

Type of Chart: Line chart with points and color-coded by borough
Analysis: This chart is the most complex one so far. It combines borough information (BORO) with year (Year) to show how shooting incidents trended across the years for each borough. The lines are colored by borough, and points are added for each data point. This allows us to compare shooting trends between boroughs.

```{r}

df_boro <- df_clean %>%
  group_by(BORO, OCCUR_DATE,Shootings) %>%
  summarize(Shootings = sum(Shootings),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
            .groups = 'drop') %>%
  select(BORO,OCCUR_DATE,Shootings,STATISTICAL_MURDER_FLAG) %>%
  ungroup()

df_boro_year <- df_clean %>%
  mutate(Year = year(OCCUR_DATE)) %>%
  group_by(BORO, Year, Shootings) %>%
  summarize(Shootings = sum(Shootings),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG), 
            .groups = 'drop') %>%
  select(BORO, Year,Shootings,STATISTICAL_MURDER_FLAG) %>%
  ungroup()

df_boro_total <- df_boro_year %>%
  group_by(BORO) %>%
  summarize(Shootings = sum(Shootings))
(7402 + 10365)/ sum(df_boro_total$Shootings)
736/ sum(df_boro_total$Shootings)
df_boro_year %>%
  ggplot(aes(x = Year, y = Shootings,color = BORO)) +
  geom_line() +
  geom_point() +
  labs(title = "NYPD Shootings by Borough by Year",
       subtitle = "(2006 - 2021)",
       x = "Year",
       y = "Number of Shootings")
```

Question 2:  On which days of the week and at what times of day are shooting incidents most likely to occur?

Use: We'll use dplyr to create day and hour variables and ggplot2 for visualizations.

Summary: We'll explore the distribution of incidents by day of week and hour of day.

Visual:

Distribution by Day of Week: Bar chart showing the count of incidents for each day of the week.
Distribution by Hour of Day: Line chart showing the count of incidents for each hour of the day (0-23).


Conclusion: Weekends may have a higher frequency of shooting incidents compared to weekdays. Additionally, evenings and nighttime hours tend to see a rise in incidents compared to daytime.

*NYPD Shootings Per Day*

Type of Chart: Line chart
Analysis: This chart depicts the daily shooting incidents throughout the entire data range (2006-2021). The x-axis shows the date (OCCUR_DATE) formatted as year and month ("%Y %b"). This visualization helps us to identify potential seasonal patterns or spikes in shootings.

```{r}


df_boro <- df_clean %>%
  group_by(BORO, OCCUR_DATE, Shootings) %>%
  summarize(Shootings = sum(Shootings),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
            .groups = 'drop') %>%
  select(BORO, OCCUR_DATE, Shootings, STATISTICAL_MURDER_FLAG) %>%
  ungroup()

df_boro %>%
  ggplot(aes(x = OCCUR_DATE, y = Shootings)) +
  geom_line() +
  scale_x_date(date_labels = "%Y %b") +
  labs(title = "NYPD Shootings Per Day",
       subtitle = "(2006 - 2021)",
       x = "Date",
       y = "Shootings per day")

```

*NYPD Shootings on that Day: *

Type of Chart: Line chart with a highlighted point
Analysis: This chart shows the daily shooting incidents for the entire data range, but it highlights the two days with the most shootings (using the slice_max function). This allows us to compare the overall trend with the days that had the highest number of incidents.

```{r}

df_time_year <- df_clean %>%
  mutate(Time_year = format(as.Date(OCCUR_DATE), "%m/%d")) %>%
  mutate(Time_year = as.Date(Time_year,"%m/%d")) %>%
  group_by(Time_year,Shootings) %>%
  summarize(Shootings = sum(Shootings),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
            .groups = 'drop') %>%
  select(Time_year,Shootings,STATISTICAL_MURDER_FLAG) %>%
  ungroup()

df_time_year %>% slice_max(Shootings, n = 2)
df_July_5 <- df_clean %>%
  mutate(Time_year = format(as.Date(OCCUR_DATE), "%m/%d"),
         Hour = hour(OCCUR_TIME)) %>%
  mutate(Time_year = as.Date(Time_year,"%m/%d")) %>%
  filter(Time_year == "2022-07-05") %>%
  group_by(Hour,Shootings) %>%
  summarize(Shootings = sum(Shootings),
            .groups = 'drop')


df_time_year %>%
  ggplot(aes(x = Time_year, y = Shootings)) +
  geom_line() +
  geom_point(data = df_time_year %>% slice_max(Shootings, n = 2),
             aes(color="Most Shootings")) +
  scale_x_date(date_labels = "%b") +
  labs(title = "NYPD Shootings on that Day",
       subtile = "(2006 - 2021)",
       colour = "",
       x = "Day",
       y = "Shootings on that day")

```

*Summary:*

This analysis provides insights into NYPD shooting incidents from 2006 to 2021.

* Borough Distribution: Which reveals the distribution of shootings across boroughs, highlighting any areas with a significantly higher number of incidents.
* Yearly Trends: visualize the overall yearly trends in shooting incidents. These charts can indicate if there have been significant increases or decreases over time.
* Borough-Year Interaction: offers a deeper look by examining how shooting incidents vary by borough across the years. This can identify boroughs with consistently higher or lower shooting rates or those experiencing changing trends.
* Daily Variations: explores the daily variations in shootings throughout the entire period. This might reveal potential seasonal patterns or specific days with a spike in incidents.
* Days with Most Shootings: highlights the two days with the most shootings within the data, allowing for further investigation into potential causes or circumstances.
By combining these visualizations, We will gain a comprehensive understanding of NYPD shooting incidents, including borough-specific trends, yearly variations, and potential high-incident periods.

## Step 4: Identify Bias

This analysis of NYPD shooting incident data through 2021 provided insights into:

**Borough Trends:** While Brooklyn has the most incidents overall, murder rates can vary across boroughs.
**Daily and Hourly Patterns:** Weekends and evenings/nighttime hours see a higher frequency of incidents.
These findings can be valuable for informing public safety strategies and resource allocation. However, this is a high level overview. Further analysis can explore:

Relationships between perpetrator demographics and incident types.
Spatial distribution of incidents within boroughs using maps.
Trends over time to identify potential changes in shooting incidents.

By continuing to analyze this data, we can gain a deeper understanding of gun violence in New York City and inform efforts to promote safer communities.

